Towards Accelerating Deep Neural Network Training on FPGAs: Facilitating the Use of Variable Precision\\ \Large{(Unrefereed Workshop Manuscript)} \\ \large{This version is amended from it's original version submitted to SWoPP 2018}

### 2018-11-27 Final Extra Project Paper/paper.tex
Added modified paper and final code for framework.

\title{Towards Accelerating Deep Neural Network Training on FPGAs: Facilitating the Use of Variable Precision\\ \Large{(Unrefereed Workshop Manuscript)}}
\affiliate{TiTech}{Tokyo Institute of Technology, 
Meguro-ku, Tokyo 152--8550, Japan}
\affiliate{TUDelft}{Delft University of Technology, Mekelweg 2, 2628 CD Delft, Netherlands}
\affiliate{Riken}{RIKEN Center for Computational Science, Chuo-ku, Kobe, Hyogo 650-0047 Japan}
\affiliate{AIST}{AIST - Tokyo Tech Real World Big-Data Computation Open Innovation Laboratory, Meguro-ku, Tokyo 152-8550, Japan}{%National Institute of Advanced Industrial Science and Technology, Chiyoda-ku, Tokyo 100-8921 Japan}

@@

The remainder of our paper is structure in the following way: \cref{sec:relwork} positions our work against other work, similar efforts.
\cref{sec:framework} gives a background to machine learning and FPGAs, and describes our proposed architecture.
\cref{sec:method} overviews our experimental methodology, and is followed by our preliminary results in \cref{sec:result}.
We conclude in \cref{sec:conclusion}.

@@

The present paper focuses on the IEEE-754 part of inference, where training and alternative data-types are on-going and future work.
Although we do have a posit~\cite{podobasposit} generator ready to be used, it has yet to be integrated into our flow.

\subsubsection{Design overview}
\Cref{fig:block-diagram} shows an overview of the system.
The backwards kernels are a work in progress, so for now we leverage external frameworks for training.
The backwards kernels are the kernels that calculate the error and back propagate the error while updating the weights and biases in the network used during training.
The host code -- written in Python -- does one forward kernel call for every layer during inference.
Finally it does a final softmax kernel call to get the final prediction.
For performance this could obviously be done on the host, given the limited size of most final layers, but performance is not the main goal of this effort.
The main goal is to simplify changing the numerical format, and thus all the calculations need to be done in that format, and an FPGA is better at accommodating this without a huge performance impact. % TODO add citation for "HUGE"

@@

The OpenCL SDK from Intel used for this design tries to pipeline loops in single work item kernels, as opposed to NDRange kernels (that are more truthful to the SIMT paradigm).
To get maximum performance the aim is to get an initiation interval of one.
That is to say that a new loop iteration is scheduled every clock cycle.
To get this to work on the inner most loop -- one that is essentially a reduction -- an array is used to store the output of every iteration in separate locations, which enables the loop to be fully unrolled.
For the loops surrounding the inner most loop, serial data-dependencies are to be avoided whenever possible, so different loop iterations are fully data-independent.
For the outermost loop that processes one block this behavior is not possible, since the output block is written to multiple times.
Multiple different blocks can theoretically be processed on the FPGA should space allow for it.
Future revisions of this framework will also include a training phase, and we thus postponed further parallelization efforts until the kernel is complete.

\subsubsection{Design decisions}
One of the main choices with the largest impact on implementation is whenever to keep all weight matrices in memory transposed.
During inference this is very beneficial, as it makes most reads sequential.
In early stages of development it sped up the kernel performance by a factor of \num{3}.
During training, using stochastic gradient descent, this can be a problem though, since the weight matrix should now be transposed back for the most efficient memory access patterns.
But since the arithmetic density of the training process is higher, the effect might be slightly less pronounced.
Another decision was to execute the outermost loop, the one that loops through all the layers, to the host.
This loop cannot be pipelined due to a serial data-dependency, so it makes more sense to execute this loop on the host CPU.
This saves space on the FPGA and simplifies the hardware design.

@@

\section{Results}\label{sec:result}
The host (NumPy based) code benchmarked at \num{60} GFLOPS for straight-forward inference of the first network and about \num{160} GFLOPS for the second network.
A first naive implementation on the FPGA clocked in at about \num{2.5} GFLOPS for this problem size.
Transposing the weight matrix sped up the kernel to around \num{7.5} GFLOPS.\@
The results for the different final FPGA kernels are shown in \cref{tab:kernel-performance}.

\begin{table}[h]
  \centering
  \caption{FPGA Kernel performance}\label{tab:kernel-performance}
  \begin{tabular}{lSS}
    \toprule
    \textbf{Block size} & \textbf{GFLOPS MNIST}  & \textbf{GFLOPS CIFAR-10}  \\
    \midrule
    64$\times$64 & 17.8 & 18.9  \\ % kernel_nn_blocking_B64inorder_hw_2018-06-28T22.52.27.aocx Network MNIST: [28 * 28, 512, 512, 10] CIFAR: [32 * 32 * 3, 1024, 512, 512, 10]
    16$\times$16 & 4.3 & 4.4 \\ % kernel_nn_blocking_B16inorder_hw_2018-06-28T22.53.27.aocx Network MNIST: [28 * 28, 512, 512, 10] CIFAR: [32 * 32 * 3, 1024, 512, 512, 10]
    \bottomrule
  \end{tabular}
\end{table}

The performance is memory-bound for these kernels. However, given our experience with alternative numerical formats (most notably posits~\cite{podobasposit}), we do not expect performance to be degraded when changing numerical format.

\subsection{FPGA Resource Utilization}
The kernel resource utilization is shown in \cref{tab:fpga-util}.
The main limiting factor is the available block RAM.
To stop the calculation of extra padding lines in the matrices, the block size should be a multiple of most of the layer sizes.
Larger block sizes would exhaust the block RAM on the used FPGA device.
The next logical size up from \num{64} would be \num{84} (\num{3}\(\times\)\num{28}) would use to much of the block RAM already.
This is because there are three block buffers and the relationship between block size and used block RAM is square.

\begin{table}[h]
  \centering
  \caption{FPGA Hardware utilization for different block sizes}\label{tab:fpga-util}
  \begin{tabular}{lllll}
    \toprule
    \textbf{Block size} & \textbf{Logic} & \textbf{DSPs} & \textbf{RAMs} & \textbf{Fmax} \\
    \midrule    
    64$\times$64 & \SI{33}{\percent} & \SI{10}{\percent} & \SI{68}{\percent} & \SI{172.8}{\mega\hertz}  \\ % kernel_nn_blocking_B64inorder_hw_2018-06-28T22.52.27.aocx
    16$\times$16 & \SI{17}{\percent} & \SI{3}{\percent} & \SI{30}{\percent} & \SI{209.16}{\mega\hertz}  \\ % kernel_nn_blocking_B16inorder_hw_2018-06-28T22.53.27.aocx
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Inferrence Performance}
The MNIST network test set accuracy after training on the CPU using Keras is \SI{97.9}{\percent}.
And the CIFAR-10 accuracy after training is \SI{47.9}{\percent}.
The FPGA gets slightly different floating point results due to rounding differences but is matches the accuracy.

\section{Conclusion}\label{sec:conclusion}
We have introduced and described our efforts to create a general-purpose deep-learning framework for design space exploration of numerical formats.
While still in the early stages, we have focused on generality.
We have shown interoperability with Keras, where we can train weights offline and import them into our framework with little effort.
We evaluated the performance and resource consumption of our design on the FPGAs, and made sure that the inferred accuracy is correct.
We are currently working on implementing the training phase, as well as the choice of numerical format to be used in both the inference and training part of our framework.

>>>>

\title{Towards Accelerating Deep Neural Network Training on FPGAs: Facilitating the Use of Variable Precision\\ \Large{(Unrefereed Workshop Manuscript)} \\ \large{This version is amended from it's original version submitted to SWoPP 2018}}
\affiliate{TiTech}{Tokyo Institute of Technology, 
Meguro-ku, Tokyo 152--8550, Japan}
\affiliate{TUDelft}{Delft University of Technology, Mekelweg 2, 2628 CD Delft, Netherlands}
\affiliate{Riken}{RIKEN Center for Computational Science, Chuo-ku, Kobe, Hyogo 650-0047 Japan}
\affiliate{AIST}{AIST - Tokyo Tech Real World Big-Data Computation Open Innovation Laboratory, Meguro-ku, Tokyo 152-8550, Japan}{%National Institute of Advanced Industrial Science and Technology, Chiyoda-ku, Tokyo 100-8921 Japan}

@@

The remainder of our paper is structure in the following way: \cref{sec:relwork} positions our work against other work, similar efforts.
\cref{sec:framework} gives a background to machine learning and FPGAs, and describes our proposed architecture.
\cref{sec:method} overviews our experimental methodology, and is followed by our results in \cref{sec:result}.
We conclude in \cref{sec:conclusion}.

@@

The present paper focuses on the IEEE-754 part of inference and training, where alternative data-types are on-going and future work.
Although we do have a posit~\cite{podobasposit} generator ready to be used, it has yet to be integrated into our flow.

\subsubsection{Design overview}
\Cref{fig:block-diagram} shows an overview of the system.
The forward kernels are the kernels the run to calculate the current activation and prediction values.
The backwards kernels are the kernels that calculate the error and back propagate the error while updating the weights and biases in the network used during training.
The host code -- written in Python -- does one forward kernel call for every layer during inference.
Finally it does a final softmax kernel call to get the final prediction.
After which it runs the first delta kernel and continues on to run the backward kernel for all layers.
For performance this could obviously be done on the host, given the limited size of most final layers, but performance is not the main goal of this effort.
The main goal is to simplify changing the numerical format, and thus all the calculations need to be done in that format, and an FPGA is better at accommodating this without a huge performance impact. % TODO add citation for "HUGE"

@@

The OpenCL SDK from Intel used for this design tries to pipeline loops in single work item kernels, as opposed to NDRange kernels (that are more truthful to the SIMT paradigm).
To get maximum performance the aim is to get an initiation interval of one.
That is to say that a new loop iteration is scheduled every clock cycle.
To get this to work on the inner most loop -- one that is essentially a reduction -- an array is used to store the output of every iteration in separate locations, which enables the loop to be fully unrolled.
For the loops surrounding the inner most loop, serial data-dependencies are to be avoided whenever possible, so different loop iterations are fully data-independent.
For the outermost loop that processes one block this behavior is not possible, since the output block is written to multiple times.
Multiple different blocks can theoretically be processed on the FPGA should space allow for it.
But due to the space required for the training logic, on our current hardware (\cref{tab:test-hardware}) this does not fit or at least it will not pass the optimization stage on the OpenCl SDK compiler.

The training logic itself, requires two matrix-matrix multiplications and three more simple multiply substract block.
Due to the size requirements and the fact that the compiler will not successfully optimize the larger designs, the blocksize of the whole design is now limited to 24 and implementing a similar acceleration as above is impossible without using vendor IP.

\subsubsection{Design decisions}
One of the main choices with the largest impact on implementation is whenever to keep all weight matrices in memory transposed.
During inference this is very beneficial, as it makes most reads sequential.
In early stages of development it sped up the kernel performance by a factor of \num{3}.
During training, using stochastic gradient descent, this can be a problem though, since the weight matrix should now be transposed back for the most efficient memory access patterns.
But since the arithmetic density of the training process is higher, the effect might be slightly less pronounced.
Although it also increases the used area slightly because now the matrix multiplication block is distinct from the other ones.
The transposing or indexing logic is different for each.
Another decision was move the execution of the outermost loop, the one that loops through all the layers, to the host.
This loop cannot be pipelined due to a serial data-dependency, so it makes more sense to execute this loop on the host CPU.
This saves space on the FPGA and simplifies the hardware design.

@@

\section{Results}\label{sec:result}
The host (NumPy based) code benchmarked at \num{60} GFLOPS for straight-forward inference of the first network and about \num{160} GFLOPS for the second network.
For the backward implementation the host code runs at \numrange{500}{800} GFLOPS, due to the higher arithmetic density.
A first naive implementation on the FPGA clocked in at about \num{2.5} GFLOPS for this problem size.
Transposing the weight matrix sped up the kernel to around \num{7.5} GFLOPS.\@
The results for the different final FPGA block sizes are shown in \cref{tab:inference-kernel-performance} for the inference kernels.
\Cref{tab:kernel-performance} shows the performance of the training kernels (they include an inference kernel), here it is clear that the training kernels are limiting the overall throughput.
The backwards kernels have a lot of potential to be optimized if required, but much is dependant on the other data types and their implementation (size), so this is better done when actually implementing this flow for that data type.

\begin{table}[h]
  \centering
  \caption{FPGA Inference Kernel performance}\label{tab:inference-kernel-performance}
  \begin{tabular}{lSS}
    \toprule
    \textbf{Block size} & \textbf{GFLOPS MNIST}  & \textbf{GFLOPS CIFAR-10}  \\
    \midrule
    64$\times$64 & 17.8 & 18.9  \\ % kernel_nn_blocking_B64inorder_hw_2018-06-28T22.52.27.aocx Network MNIST: [28 * 28, 512, 512, 10] CIFAR: [32 * 32 * 3, 1024, 512, 512, 10]
    16$\times$16 & 4.3 & 4.4 \\ % kernel_nn_blocking_fixed2BS16_hw_2018-11-08T05.26.51.aocx Network MNIST: [28 * 28, 512, 512, 10] CIFAR: [32 * 32 * 3, 1024, 512, 512, 10]
    \bottomrule
  \end{tabular}
\end{table}

\begin{table}[h]
  \centering
  \caption{FPGA Full Kernel performance}\label{tab:kernel-performance}
  \begin{tabular}{lSS}
    \toprule
    \textbf{Block size} & \textbf{GFLOPS MNIST}  & \textbf{GFLOPS CIFAR-10}  \\
    \midrule
    24$\times$24 & 6.6 & 6.8  \\ % kernel_nn_blocking_fixed2BS24_hw_2018-11-08T05.20.20.aocx Network MNIST: [28 * 28, 512, 512, 10] CIFAR: [32 * 32 * 3, 1024, 512, 512, 10]
    16$\times$16 & 6.3 & 6.4 \\ % kernel_nn_blocking_fixed2BS16_hw_2018-11-08T05.26.51.aocx Network MNIST: [28 * 28, 512, 512, 10] CIFAR: [32 * 32 * 3, 1024, 512, 512, 10]
    \bottomrule
  \end{tabular}
\end{table}

The performance is mostly memory-bound for these kernels. 
Especially the backwards kernels seem to have a hard limit at around \numrange{6}{7} GFLOPS in their current form.
Furthermore, given our experience with alternative numerical formats (most notably posits~\cite{podobasposit}), we do not expect performance to be degraded when changing numerical format.

\subsection{FPGA Resource Utilization}
The kernel resource utilization is shown in \cref{tab:fpga-inference-util} for the inference kernels only and \cref{tab:fpga-util} shows the utilization for the full implementation.
The main limiting factor is the available block RAM and the ability for the compiler to runs its optimization phase without crashing.
To stop the calculation of extra padding lines in the matrices, the block size should be a multiple of most of the layer sizes.
Larger block sizes would exhaust the block RAM on the used FPGA device.
The next logical size up from \num{64} would be \num{84} (\num{3}\(\times\)\num{28}) would use to much of the block RAM already for the inference alone.
This is because there are three block buffers and the relationship between block size and used block RAM is square.
With the training kernels the theoretical need for block RAMs would triple if all matrix operations were implemented in the same way.
Right now on a single device, make the limit to the block size rather small, in light of that, it was decided to postpone the further optimizations after different formats were added.
Those will also take a much larger area themselves, making the problem worse.

\begin{table}[h]
  \centering
  \caption{FPGA Inference Hardware utilization for different block sizes}\label{tab:fpga-inference-util}
  \begin{tabular}{lllll}
    \toprule
    \textbf{Block size} & \textbf{Logic} & \textbf{DSPs} & \textbf{RAMs} & \textbf{Fmax} \\
    \midrule    
    64$\times$64 & \SI{33}{\percent} & \SI{10}{\percent} & \SI{68}{\percent} & \SI{172.8}{\mega\hertz}  \\ % kernel_nn_blocking_B64inorder_hw_2018-06-28T22.52.27.aocx
    16$\times$16 & \SI{17}{\percent} & \SI{3}{\percent} & \SI{30}{\percent} & \SI{209.16}{\mega\hertz}  \\ % kernel_nn_blocking_B16inorder_hw_2018-06-28T22.53.27.aocx
    \bottomrule
  \end{tabular}
\end{table}

\begin{table}[h]
  \centering
  \caption{FPGA Full Hardware utilization for different block sizes}\label{tab:fpga-util}
  \begin{tabular}{lllll}
    \toprule
    \textbf{Block size} & \textbf{Logic} & \textbf{DSPs} & \textbf{RAMs} & \textbf{Fmax} \\
    \midrule    
    24$\times$24 & \SI{30}{\percent} & \SI{7}{\percent} & \SI{50}{\percent} & \SI{223.51}{\mega\hertz}  \\ % kernel_nn_blocking_fixed2BS24_hw_2018-11-08T05.20.20.aocx
    16$\times$16 & \SI{23}{\percent} & \SI{6}{\percent} & \SI{39}{\percent} & \SI{218.24}{\mega\hertz}  \\ % kernel_nn_blocking_fixed2BS16_hw_2018-11-08T05.26.51.aocx
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Inference Performance}
The MNIST network test set accuracy after training on the CPU is \SI{97.9}{\percent}.
And the CIFAR-10 accuracy after training is \SI{47.9}{\percent}.
The FPGA gets slightly different floating point results due to rounding differences and reaches \SI{95.4}{\percent} and \SI{45.2}{\percent} respectively after the same number of epochs.

\section{Conclusion}\label{sec:conclusion}
We have introduced and described our efforts to create a general-purpose deep-learning framework for design space exploration of numerical formats.
While still in the early stages, we have focused on generality.
We have shown interoperability with Keras, where we can train weights offline and import them into our framework with little effort.
Also that we can correctly train on the CPU and FPGA with custom implementations of stochastic gradient descent.
We evaluated the performance and resource consumption of our design on the FPGAs, and made sure that the inferred accuracy is correct.
And that the training performance sufficient for further testing.
We are currently working on the choice of numerical format to be used in both the inference and training part of our framework.
